Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-04-26 paper/ms.tex
Replace N-body with macro

@@

\begin{figure*}[h!]%
\centering
\includegraphics[width=0.8\textwidth]{fig/fig1.pdf}
\caption{Surface density projections in an N-body only simulation (upper
panels) and a simulation which includes the SMUGGLE model (bottom panels). We
can see that in the N-body run, the bar grows in length and strength. In the
SMUGGLE run (which includes a gaseous phase and a model for the multi-phase
interstellar medium), the bar remains at approximately the same length and
strength over the course of the simulation. The N-body model is identical to
the GALAKOS model, discussed in the text.}\label{fig:overview}
\end{figure*}

\begin{figure}[h!]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig2.pdf}
\caption{Various bar properties from the N-body and SMUGGLE runs. The
\textit{upper panel} show the evolution of the pattern speed. As expected, the
bar in the N-body run slows down due to interactions between the bar and the
dark matter halo. However, the bar in the SMUGGLE run does not slow down and
instead remains at a constant pattern speed. The \textit{middle panel} shows
the evolution of the bar length. In the N-body case, the bar lengthens. This
occurs because as the pattern speed drops, bar-like orbits at larger radii are
possible. Stars are captured on these orbits, lengthening the bar. This
process does not occur in the SMUGGLE cases since the bar pattern speed is not
decreasing, and therefore the bar length remains constant. The \textit{lower
panel} shows the torque on the bar by different components. The solid lines
show the torque exerted by the halo in both the N-body and SMUGGLE cases. The
dashed line shows the torque exerted by the gas phase in the SMUGGLE run
(there is no gas in the N-body run). Details on how these properties
are calculated is given in the Methods section.}\label{fig:prop}
\end{figure}

\begin{figure*}[h!]%
\centering
\includegraphics[width=0.9\textwidth]{fig/fig3.pdf}
\caption{The wake excited in the dark matter halo in the N-body case
(\textit{left panel}) and SMUGGLE case (\textit{right panel}) after
$2.6\,\textrm{Gyr}$. The \textit{left} and \textit{right panels} show a
surface density projection in the $x$-$y$ plane of the dark matter halo after
an axisymmetric average has been subracted. The solid line indicates the
direction of the bar while the dashed line indicates the direction of the halo
wake (both measured by taking the second Fourier component within a sphere of
all material within a radius of $X\,\textrm{kpc}$). The \textit{center panel}
shows the time evolution of the angle difference between the bar and the halo
wake, as measured from the second Fourier component. After the first Gyr, the
angle difference in the SMUGGLE case is smaller than in the N-body case by
about a factor of two, reflecting how the dark matter halo in the SMUGGLE case
is unable to exert as negative a torque on the bar as in the N-body
case.}\label{fig:wake}
\end{figure*}

\begin{figure}[h!]
\centering
\includegraphics{fig/fig4.pdf}
\caption{The torque on the bar by the gas when the pattern speed of the bar is
kept at a constant value. Only gas within the corotation radius is able to
infall. Since slower bars have larger corotation radii, slower bars experience
a larger net torque than faster bars. The setup of the simulations used here
is identical to the SMUGGLE case discussed earlier and in the Methods section,
except the N-body disk is rotated as a solid body with a constant angular
velocity.}\label{fig:equil}
\end{figure}

@@

A surface density projection of our simulations is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the N-body
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large
qualitative difference in the evolution of the bar pattern between the two
runs. We see that in the N-body case, the bar lengthens in time and grows in
strength. In the SMUGGLE case, the bar retains a similar length and strength
between panels.

We show the time evolution of different bar properties in Fig.~\ref{fig:prop}.
In the upper panel, we show the pattern speed over time in the N-body (blue)
and SMUGGLE (orange) runs. The pattern speed in the N-body case slows down
while the pattern speed in the SMUGGLE case remains roughly constant. The
slowing down of the pattern speed in the N-body case is consistent with a long
line of numerical research on bars in N-body
simulations.\cite{1992ApJ...400...80H, 2000ApJ...543..704D,
2002MNRAS.330...35A, 2002ApJ...569L..83A, 2003MNRAS.341.1179A,
2003MNRAS.346..251O, 2005MNRAS.363..991H, 2006ApJ...637..214M,
2007MNRAS.375..460W, 2009ApJ...697..293D} However, in the SMUGGLE case the
pattern speed remains constant. After the first Gyr of evolution, we find that
the pattern speed increases by only $\sim10\%$ over the next
$4\,\textrm{Gyr}$, compared to a $\sim43\%$ decrease in the pattern speed for
the N-body run over the same interval. As we saw qualitatively in
Fig.~\ref{fig:overview}, the length of the bar in the N-body case grows over
time while it remains roughly constant in the SMUGGLE case. This is also
consistent with previous numerical work, which found that bars tend to grow as
they slow down and the radius of corotation
increases.\cite{2000ApJ...543..704D, 2003MNRAS.341.1179A}

The bottom panel of Fig.~\ref{fig:prop} shows the torque exerted on the bar by
different components. The solid lines indicate the torque on the bar by the
dark matter halo whereas the dashed line indicates the torque on the bar by
the gas phase. In the N-body case, the halo exerts a steady negative torque on
the bar, with an average torque from $1$ to $4\,\textrm{Gyr}$ of $-58.0$ in
units of $10^{10}M_{\odot}\,(\textrm{km}/\textrm{s})^2$. The halo in the
SMUGGLE case exerts a similar negative torque on the bar in the first Gyr of
evolution, but after that the halo exerts a much smaller torque on the bar,
averaging only $-7.8$ in the same units and over the same interval. The gas in
the SMUGGLE case exerts a steady positive torque averaging $11.7$ in the same
units and over the same interval.

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
positive torque on the bar can be understood in terms of the halo wake
mechanism. In the N-body case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
which slows it down.\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to participate in forming the wake. However, the
gas is a reliable source of positive torque on the bar, speeding the bar up.
This stops the resonance location from changing such that the halo cannot
reinforce a wake, arresting the process by which the halo can slow the bar
down.

We can test this interpretation by measuring the angle offset between the halo
wake and the bar. If the wake and the bar are aligned (i.e., there is no angle
offset), then the wake cannot exert a negative torque on the bar. This angle
is plotted in the middle panel of Fig.~\ref{fig:wake}, which shows that the
angle offset is larger in the N-body case than in the SMUGGLE case by about a
factor of two. The left and right panels of Fig.~\ref{fig:wake} show the halo
wake with respect to the location of the bar in the N-body (left) and SMUGGLE
(right) cases.

@@

Second, we performed a set of simulations in which we decreased the initial
gas mass of the disk. We essentially find that there is a certain critical gas
fraction (maybe gas surface density?) above which our proposed mechanism
operates. If the gas available is below this threshold, then the pattern speed
will decrease in a similar fashion to standard N-body simulations.
Importantly, arbitrarily increasing the gas fraction does not cause the
pattern speed to continuously increase. This is consistent with our proposed
equilibrium mechanism and with previous studies of barred galaxies which
varied the gas fraction.\cite{2010ApJ...719.1470V} The exact critical gas
fraction above which our mechanism operates is likely dependent on the
properties of the bar, disk, and dark matter halo -- further work is needed to
characterize how this gas fraction varies with such properties. Details on
this test in which we varied the gas mass of the disk are given in the Methods
section.

@@

\noindent
{\bf Initial Setup}
\\
\noindent
The initial setup of the galactic disk used in this work follows closely the
GALAKOS model\cite{2020ApJ...890..117D}, which uses a modified version of
\texttt{MakeNewDisk}.\cite{2005MNRAS.361..776S} The GALAKOS model has three
components - a radially exponential and vertically isothermal stellar disk,
and a stellar bulge and dark matter halo following a Hernquist
profile.\cite{1990ApJ...356..359H} All N-body runs in this work used the same
setup parameters as the GALAKOS disk, more details of which can be found in
the original paper.

@@

After generating the gaseous disk in this way, we stitched the gas disk
together with the GALAKOS N-body disk (and bulge and dark matter halo) after
the GALAKOS disk has been allowed to evolve for $1.5\,\textrm{Gyr}$. The
purpose of allowing the GALAKOS disk to evolve first for a short period of
time is to allow for the bar to form unimpacted by the presence of the gas
(which would normally disrupt the formation of the bar). Throughout this work,
we consider $t=0$ for the N-body run to be the time at which we added the gas
phase for the SMUGGLE run (i.e. we ignore the first $1.5\,\textrm{Gyr}$ of
evolution of the N-body disk when the bar is forming). We made one additional
modification when stitching the gas disk together with the N-body disk - we
created a hole within the central $4\,\textrm{kpc}$. This hole guards
against an initial dramatic infall of gas within the bar region, which we
found to destroy the bar.

\begin{figure}[t]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-vcirc.pdf}
\caption{The circular velocity curve for the N-body run (blue) and the SMUGGLE
run (orange) compared to observational estimates.\cite{2019ApJ...871..120E} We
see that the circular velocity curve for both runs is marginally larger than
the Milky Way's, but still comparable. The SMUGGLE circular velocity curve is
larger than the N-body curve due to the additional mass in the gas phase.}
\label{fig:vcirc}
\end{figure}

@@

\begin{figure}[t!]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-A2.pdf}
\caption{The time evolution of the bar strength, measured as the second
Fourier component divided by the zeroth Fourier component (a formula is given
in the Methods section). We see that in the N-body case (blue) the bar
strength increases with time, consistent with previous results showing that
the bar strength increases as bars slow down. In the SMUGGLE case (orange), we
see that the bar strength remains roughly constant, possibly slightly
decreasing with time. This is also consistent with the expected relation
between pattern speed and strength since the bar in this case is not slowing
down.}
\label{fig:strength}
\end{figure}

\begin{figure*}[t!]%
\centering
\includegraphics[width=0.95\textwidth]{fig/fig-bar_decomp.pdf}
\caption{Demonstration of our bar decomposition procedure based on
ref.\cite{2016MNRAS.463.1952P} The \textit{left panel} shows a surface density
projection through the stellar component of the N-body simulation (disk and
bulge). The \textit{middle panel} shows the component of the disk identified
as being trapped in the bar while the \textit{right panel} shows the component
of the disk identified as not being trapped in the bar. The fact that the
untrapped stars form a roughly axisymmetric structure indicates our bar
decomposition is sufficiently accurate. {\color{red} I will remake this figure
- I ended up losing the files which stored the bar membership, so I just
pulled these screenshots from an earlier presentation.}}
\label{fig:decomp}
\end{figure*}

The time evolution of the bar strength, defined as the maximum of
$\left|A_2/A_0\right|$ as a function of radius, is shown in Extended Data
Fig.~\ref{fig:strength}. The quantity $\left|A_2/A_0\right|$ varies from $0$
to $1$, with larger values indicating a stronger bar pattern. We see that in
the N-body case, $\left|A_2/A_0\right|$ increases over time as the bar pattern
slows. This is consistent with previous N-body simulations which showed a
clear correlation between the bar pattern speed and the bar
strength.\cite{2003MNRAS.341.1179A} In the SMUGGLE case, we see that the bar
strength has an initial drop but then remains at a roughly constant but
slightly decreasing strength. This is consistent with the pattern speed in the
SMUGGLE case being roughly constant or slightly increasing.

>>>>

\newcommand{\Nbody}{$N$-body}

@@

\begin{figure*}[h!]%
\centering
\includegraphics[width=0.8\textwidth]{fig/fig1.pdf}
\caption{Surface density projections in an \Nbody{} only simulation (upper
panels) and a simulation which includes the SMUGGLE model (bottom panels). We
can see that in the \Nbody{} run, the bar grows in length and strength. In the
SMUGGLE run (which includes a gaseous phase and a model for the multi-phase
interstellar medium), the bar remains at approximately the same length and
strength over the course of the simulation. The \Nbody{} model is identical to
the GALAKOS model, discussed in the text.}\label{fig:overview}
\end{figure*}

\begin{figure}[h!]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig2.pdf}
\caption{Various bar properties from the \Nbody{} and SMUGGLE runs. The
\textit{upper panel} show the evolution of the pattern speed. As expected, the
bar in the \Nbody{} run slows down due to interactions between the bar and the
dark matter halo. However, the bar in the SMUGGLE run does not slow down and
instead remains at a constant pattern speed. The \textit{middle panel} shows
the evolution of the bar length. In the \Nbody{} case, the bar lengthens. This
occurs because as the pattern speed drops, bar-like orbits at larger radii are
possible. Stars are captured on these orbits, lengthening the bar. This
process does not occur in the SMUGGLE cases since the bar pattern speed is not
decreasing, and therefore the bar length remains constant. The \textit{lower
panel} shows the torque on the bar by different components. The solid lines
show the torque exerted by the halo in both the \Nbody{} and SMUGGLE cases. The
dashed line shows the torque exerted by the gas phase in the SMUGGLE run
(there is no gas in the \Nbody{} run). Details on how these properties
are calculated is given in the Methods section.}\label{fig:prop}
\end{figure}

\begin{figure*}[h!]%
\centering
\includegraphics[width=0.9\textwidth]{fig/fig3.pdf}
\caption{The wake excited in the dark matter halo in the \Nbody{} case
(\textit{left panel}) and SMUGGLE case (\textit{right panel}) after
$2.6\,\textrm{Gyr}$. The \textit{left} and \textit{right panels} show a
surface density projection in the $x$-$y$ plane of the dark matter halo after
an axisymmetric average has been subracted. The solid line indicates the
direction of the bar while the dashed line indicates the direction of the halo
wake (both measured by taking the second Fourier component within a sphere of
all material within a radius of $X\,\textrm{kpc}$). The \textit{center panel}
shows the time evolution of the angle difference between the bar and the halo
wake, as measured from the second Fourier component. After the first Gyr, the
angle difference in the SMUGGLE case is smaller than in the \Nbody{} case by
about a factor of two, reflecting how the dark matter halo in the SMUGGLE case
is unable to exert as negative a torque on the bar as in the \Nbody{}
case.}\label{fig:wake}
\end{figure*}

\begin{figure}[h!]
\centering
\includegraphics{fig/fig4.pdf}
\caption{The torque on the bar by the gas when the pattern speed of the bar is
kept at a constant value. Only gas within the corotation radius is able to
infall. Since slower bars have larger corotation radii, slower bars experience
a larger net torque than faster bars. The setup of the simulations used here
is identical to the SMUGGLE case discussed earlier and in the Methods section,
except the \Nbody{} disk is rotated as a solid body with a constant angular
velocity.}\label{fig:equil}
\end{figure}

@@

A surface density projection of our simulations is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the \Nbody{}
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large
qualitative difference in the evolution of the bar pattern between the two
runs. We see that in the \Nbody{} case, the bar lengthens in time and grows in
strength. In the SMUGGLE case, the bar retains a similar length and strength
between panels.

We show the time evolution of different bar properties in Fig.~\ref{fig:prop}.
In the upper panel, we show the pattern speed over time in the \Nbody{} (blue)
and SMUGGLE (orange) runs. The pattern speed in the \Nbody{} case slows down
while the pattern speed in the SMUGGLE case remains roughly constant. The
slowing down of the pattern speed in the \Nbody{} case is consistent with a long
line of numerical research on bars in \Nbody{}
simulations.\cite{1992ApJ...400...80H, 2000ApJ...543..704D,
2002MNRAS.330...35A, 2002ApJ...569L..83A, 2003MNRAS.341.1179A,
2003MNRAS.346..251O, 2005MNRAS.363..991H, 2006ApJ...637..214M,
2007MNRAS.375..460W, 2009ApJ...697..293D} However, in the SMUGGLE case the
pattern speed remains constant. After the first Gyr of evolution, we find that
the pattern speed increases by only $\sim10\%$ over the next
$4\,\textrm{Gyr}$, compared to a $\sim43\%$ decrease in the pattern speed for
the \Nbody{} run over the same interval. As we saw qualitatively in
Fig.~\ref{fig:overview}, the length of the bar in the \Nbody{} case grows over
time while it remains roughly constant in the SMUGGLE case. This is also
consistent with previous numerical work, which found that bars tend to grow as
they slow down and the radius of corotation
increases.\cite{2000ApJ...543..704D, 2003MNRAS.341.1179A}

The bottom panel of Fig.~\ref{fig:prop} shows the torque exerted on the bar by
different components. The solid lines indicate the torque on the bar by the
dark matter halo whereas the dashed line indicates the torque on the bar by
the gas phase. In the \Nbody{} case, the halo exerts a steady negative torque on
the bar, with an average torque from $1$ to $4\,\textrm{Gyr}$ of $-58.0$ in
units of $10^{10}M_{\odot}\,(\textrm{km}/\textrm{s})^2$. The halo in the
SMUGGLE case exerts a similar negative torque on the bar in the first Gyr of
evolution, but after that the halo exerts a much smaller torque on the bar,
averaging only $-7.8$ in the same units and over the same interval. The gas in
the SMUGGLE case exerts a steady positive torque averaging $11.7$ in the same
units and over the same interval.

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
positive torque on the bar can be understood in terms of the halo wake
mechanism. In the \Nbody{} case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
which slows it down.\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to participate in forming the wake. However, the
gas is a reliable source of positive torque on the bar, speeding the bar up.
This stops the resonance location from changing such that the halo cannot
reinforce a wake, arresting the process by which the halo can slow the bar
down.

We can test this interpretation by measuring the angle offset between the halo
wake and the bar. If the wake and the bar are aligned (i.e., there is no angle
offset), then the wake cannot exert a negative torque on the bar. This angle
is plotted in the middle panel of Fig.~\ref{fig:wake}, which shows that the
angle offset is larger in the \Nbody{} case than in the SMUGGLE case by about a
factor of two. The left and right panels of Fig.~\ref{fig:wake} show the halo
wake with respect to the location of the bar in the \Nbody{} (left) and SMUGGLE
(right) cases.

@@

Second, we performed a set of simulations in which we decreased the initial
gas mass of the disk. We essentially find that there is a certain critical gas
fraction (maybe gas surface density?) above which our proposed mechanism
operates. If the gas available is below this threshold, then the pattern speed
will decrease in a similar fashion to standard \Nbody{} simulations.
Importantly, arbitrarily increasing the gas fraction does not cause the
pattern speed to continuously increase. This is consistent with our proposed
equilibrium mechanism and with previous studies of barred galaxies which
varied the gas fraction.\cite{2010ApJ...719.1470V} The exact critical gas
fraction above which our mechanism operates is likely dependent on the
properties of the bar, disk, and dark matter halo -- further work is needed to
characterize how this gas fraction varies with such properties. Details on
this test in which we varied the gas mass of the disk are given in the Methods
section.

@@

\noindent
{\bf Initial Setup}
\\
\noindent
The initial setup of the galactic disk used in this work follows closely the
GALAKOS model\cite{2020ApJ...890..117D}, which uses a modified version of
\texttt{MakeNewDisk}.\cite{2005MNRAS.361..776S} The GALAKOS model has three
components - a radially exponential and vertically isothermal stellar disk,
and a stellar bulge and dark matter halo following a Hernquist
profile.\cite{1990ApJ...356..359H} All \Nbody{} runs in this work used the same
setup parameters as the GALAKOS disk, more details of which can be found in
the original paper.

@@

After generating the gaseous disk in this way, we stitched the gas disk
together with the GALAKOS \Nbody{} disk (and bulge and dark matter halo) after
the GALAKOS disk has been allowed to evolve for $1.5\,\textrm{Gyr}$. The
purpose of allowing the GALAKOS disk to evolve first for a short period of
time is to allow for the bar to form unimpacted by the presence of the gas
(which would normally disrupt the formation of the bar). Throughout this work,
we consider $t=0$ for the \Nbody{} run to be the time at which we added the gas
phase for the SMUGGLE run (i.e. we ignore the first $1.5\,\textrm{Gyr}$ of
evolution of the \Nbody{} disk when the bar is forming). We made one additional
modification when stitching the gas disk together with the \Nbody{} disk - we
created a hole within the central $4\,\textrm{kpc}$. This hole guards
against an initial dramatic infall of gas within the bar region, which we
found to destroy the bar.

\begin{figure}[t]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-vcirc.pdf}
\caption{The circular velocity curve for the \Nbody{} run (blue) and the SMUGGLE
run (orange) compared to observational estimates.\cite{2019ApJ...871..120E} We
see that the circular velocity curve for both runs is marginally larger than
the Milky Way's, but still comparable. The SMUGGLE circular velocity curve is
larger than the \Nbody{} curve due to the additional mass in the gas phase.}
\label{fig:vcirc}
\end{figure}

@@

\begin{figure}[t!]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-A2.pdf}
\caption{The time evolution of the bar strength, measured as the second
Fourier component divided by the zeroth Fourier component (a formula is given
in the Methods section). We see that in the \Nbody{} case (blue) the bar
strength increases with time, consistent with previous results showing that
the bar strength increases as bars slow down. In the SMUGGLE case (orange), we
see that the bar strength remains roughly constant, possibly slightly
decreasing with time. This is also consistent with the expected relation
between pattern speed and strength since the bar in this case is not slowing
down.}
\label{fig:strength}
\end{figure}

\begin{figure*}[t!]%
\centering
\includegraphics[width=0.95\textwidth]{fig/fig-bar_decomp.pdf}
\caption{Demonstration of our bar decomposition procedure based on
ref.\cite{2016MNRAS.463.1952P} The \textit{left panel} shows a surface density
projection through the stellar component of the \Nbody{} simulation (disk and
bulge). The \textit{middle panel} shows the component of the disk identified
as being trapped in the bar while the \textit{right panel} shows the component
of the disk identified as not being trapped in the bar. The fact that the
untrapped stars form a roughly axisymmetric structure indicates our bar
decomposition is sufficiently accurate. {\color{red} I will remake this figure
- I ended up losing the files which stored the bar membership, so I just
pulled these screenshots from an earlier presentation.}}
\label{fig:decomp}
\end{figure*}

The time evolution of the bar strength, defined as the maximum of
$\left|A_2/A_0\right|$ as a function of radius, is shown in Extended Data
Fig.~\ref{fig:strength}. The quantity $\left|A_2/A_0\right|$ varies from $0$
to $1$, with larger values indicating a stronger bar pattern. We see that in
the \Nbody{} case, $\left|A_2/A_0\right|$ increases over time as the bar pattern
slows. This is consistent with previous \Nbody{} simulations which showed a
clear correlation between the bar pattern speed and the bar
strength.\cite{2003MNRAS.341.1179A} In the SMUGGLE case, we see that the bar
strength has an initial drop but then remains at a roughly constant but
slightly decreasing strength. This is consistent with the pattern speed in the
SMUGGLE case being roughly constant or slightly increasing.
